---
title: 'Sydnor et al: Ketamine and White Matter Microstructure Results'
author: "Valerie Jill Sydnor"
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    toc: yes
    toc_depth: '5'
  html_document:
    toc: yes
    toc_depth: '5'
    df_print: paged
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "./")
knitr::opts_chunk$set(comment = NA)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
library(knitr)
library(rmdformats)
library(dplyr)
library(reader)
library(kableExtra)
library(ggplot2)
library(ppcor)
library(xtable)
```

```{r}
ketamine <- read.csv("../MGHKetamine_Clinical_Imaging_Data.csv", header=TRUE )
```

# RESULTS

## Replication of Ketamine’s Antidepressant Efficacy for TRD

*Study Sample Age*
```{r}
age.mean <- mean(ketamine$Age)
age.sd <- sd(ketamine$Age)
cat(age.mean, age.sd)
```

*Study Sample Sex Distribution*
```{r}
(sum(ketamine$Gender == "0"))/(nrow(ketamine))
```

*HAM-D28 24-hour percent improvement*
```{r}
hamd.improve.mean <- mean(ketamine$HDRS.24hr.percentchange)
hamd.improve.sd <- sd(ketamine$HDRS.24hr.percentchange)
cat(hamd.improve.mean, hamd.improve.sd)
```

* Treatment Responders
```{r}
responders <- ketamine %>% filter(HDRS.24hr.percentchange > 50) #identify responders
print("Number of responders")
nrow(responders)
responders.improve.mean <- mean(responders$HDRS.24hr.percentchange)
responders.improve.sd <- sd(responders$HDRS.24hr.percentchange)
cat(responders.improve.mean, responders.improve.sd)
```  

* Treatment Non-Responders
```{r}
nonresponders <- ketamine %>% filter(HDRS.24hr.percentchange < 50) #identify non-responders
print("Number of non-responders")
nrow(nonresponders)
nonresponders.improve.mean <- mean(nonresponders$HDRS.24hr.percentchange)
nonresponders.improve.sd <- sd(nonresponders$HDRS.24hr.percentchange)
cat(nonresponders.improve.mean, nonresponders.improve.sd)
``` 

* Correlation between baseline HAM-D28 score and 24-hour HAM-D28 percent improvement
```{r}
cor.test(ketamine$Baseline_HDRS, ketamine$HDRS.24hr.percentchange)
```

\newpage

## Pre-Infusion White Matter FA is Associated with 24-Hour Depression Improvement

*Partial Correlations: 24-hour HAM-D28 Percent Improvement and Baseline FA*

```{r}
R.value <- double() #create empty vector
p.value <- double() #create empty vector

for(bundle in (c("Lifo", "Rifo", "Lcing", "Rcing", "Fmajor","Fminor","Lilf","Rilf",
                 "Lslf","Rslf","Lunc","Runc","Fornix","Lcing2","Rcing2"))){
  colname_baselineFA <- sprintf("%s_baselineFA",bundle)
  baselineFA.HAMD28change.partialcorr <-
    pcor.test(ketamine[,colname_baselineFA],
              ketamine$HDRS.24hr.percentchange, ketamine$Age) #partial correlation controlled for age 
  Rvalue <- baselineFA.HAMD28change.partialcorr$estimate #Obtain R value from partial correlation
  pvalue <- baselineFA.HAMD28change.partialcorr$p.value #Obtain p-value from partial correlation
  R.value[bundle] <- Rvalue #Save R into vector
  p.value[bundle] <- pvalue #Save p-value into vector
}

BaselineFA.HAMD28Change.PartialCorrelations <- data.frame(R.value, p.value)
BaselineFA.HAMD28Change.PartialCorrelations %>% arrange(p.value)

```

* Normality of Residuals
```{r}
par(mfrow=c(1,2))

Shapiro.p.value <- double() #create empty vector

for(bundle in (c("Lifo", "Rifo", "Lcing", "Rcing", "Fmajor","Fminor","Lilf","Rilf",
                 "Lslf","Rslf","Lunc","Runc","Fornix","Lcing2","Rcing2"))){
  colname_baselineFA <- sprintf("%s_baselineFA",bundle)
  baselineFA.HAMD28change.lm <- lm(ketamine[,colname_baselineFA] ~ ketamine$HDRS.24hr.percentchange + ketamine$Age) 
  shapiro <- shapiro.test(baselineFA.HAMD28change.lm$residuals)$p.value #Shapiro-Wilx
  Shapiro.p.value[bundle] <- shapiro 
  plot(baselineFA.HAMD28change.lm, 2) #Q-Q plot
}
```

```{r}
Shapiro.p.value
```

*Partial Correlations: Baseline HAM-D28 and Baseline FA*

* Left CB-hippocampal portion 
```{r}
pcor.test(ketamine$Baseline_HDRS, ketamine$Lcing2_baselineFA, ketamine$Age)
```

* Left SLF 
```{r}
pcor.test(ketamine$Baseline_HDRS, ketamine$Lslf_baselineFA, ketamine$Age)
```

\newpage 

## White Matter FA Increases from Pre- to Post-Infusion

*Paired T-tests: Pre- and Post- Infusion FA*
```{r}
percentchange.mean <- double() #create empty vector
t.value <- double() #create empty vector
p.value <- double() #create empty vector

for(bundle in (c("Lifo", "Rifo", "Lcing", "Rcing", "Fmajor","Fminor","Lilf","Rilf",
                 "Lslf","Rslf","Lunc","Runc","Fornix","Lcing2","Rcing2"))){
  
  colname_percentchange <- sprintf("%s_FA_percentchange",bundle)
  bundle.percentchange.mean <- mean(ketamine[,colname_percentchange])
  percentchange.mean[bundle] <- bundle.percentchange.mean

  colname_baselineFA <- sprintf("%s_baselineFA",bundle)
  colname_postinfusionFA <- sprintf("%s_postinfusionFA",bundle)
  FA.paired.ttest <- t.test(ketamine[,colname_postinfusionFA],
                           ketamine[,colname_baselineFA],
                           paired=TRUE,alternative="two.sided") #paired samples t-test
  tvalue <- FA.paired.ttest$statistic #Obtain t-value from paired samples t-test
  pvalue <- FA.paired.ttest$p.value #Obtain p-value from paired samples t-test
  t.value[bundle] <- tvalue #Save t into vector
  p.value[bundle] <- pvalue #Save p-value into vector
}

PreInfusion.PostInfusion.FA.Paired.TTests <- data.frame(percentchange.mean, t.value, p.value)
PreInfusion.PostInfusion.FA.Paired.TTests %>% arrange(desc(percentchange.mean))
```
\newpage

## Post-Ketamine Change in White Matter FA is Associated with 24-Hour Depression Improvement

*Partial Correlations: 24-hour HAM-D28 Percent Improvement and Post-ketamine % change in FA*

```{r}
R.value <- double() #create empty vector
p.value <- double() #create empty vector

for(bundle in (c("Lifo", "Rifo", "Lcing", "Rcing", "Fmajor","Fminor","Lilf","Rilf",
                 "Lslf","Rslf","Lunc","Runc","Fornix","Lcing2","Rcing2"))){
  colname_percentchange <- sprintf("%s_FA_percentchange",bundle)
  percentchangeFA.HAMD28change.partialcorr <-
    pcor.test(ketamine[,colname_percentchange],
              ketamine$HDRS.24hr.percentchange, ketamine$Age) #partial correlation controlled for age 
  Rvalue <- percentchangeFA.HAMD28change.partialcorr$estimate #Obtain R value from partial correlation
  pvalue <- percentchangeFA.HAMD28change.partialcorr$p.value #Obtain p-value from partial correlation
  R.value[bundle] <- Rvalue #Save R into vector
  p.value[bundle] <- pvalue #Save p-value into vector
}

PercentChangeFA.HAMD28Change.PartialCorrelations <- data.frame(R.value, p.value)
PercentChangeFA.HAMD28Change.PartialCorrelations %>% arrange(p.value)

```
* Normality of Residuals
```{r}
par(mfrow=c(1,2))

Shapiro.p.value <- double() #create empty vector

for(bundle in (c("Lifo", "Rifo", "Lcing", "Rcing", "Fmajor","Fminor","Lilf","Rilf",
                 "Lslf","Rslf","Lunc","Runc","Fornix","Lcing2","Rcing2"))){
  colname_percentchange <- sprintf("%s_FA_percentchange",bundle)
  percentchangeFA.HAMD28change.lm <- lm(ketamine[,colname_percentchange] ~ ketamine$HDRS.24hr.percentchange + ketamine$Age) 
  shapiro <- shapiro.test(percentchangeFA.HAMD28change.lm$residuals)$p.value #Shapiro-Wilx
  Shapiro.p.value[bundle] <- shapiro 
  plot(percentchangeFA.HAMD28change.lm,2) #Q-Q Plot
}
```

```{r}
Shapiro.p.value
```

* CC-forceps minor outlier removal: partial correlation and normality of residuals 

```{r}
fminor <- ketamine %>% filter(Fminor_FA_percentchange < 3)
pcor.test(fminor$Fminor_FA_percentchange, fminor$HDRS.24hr.percentchange, fminor$Age)
```
```{r}
fminor.percentchangeFA.HAMD28change.lm <- lm(fminor$Fminor_FA_percentchange ~ 
                                          fminor$HDRS.24hr.percentchange + fminor$Age)
plot(fminor.percentchangeFA.HAMD28change.lm,2)
shapiro.test(fminor.percentchangeFA.HAMD28change.lm$residuals)
```
\newpage 

## Sensitivity Analyses: Assessing Head Motion and High Leverage Observations 

### Head Motion

*Baseline FA and Baseline in-scanner head motion*
```{r}
p.value <- double() #create empty vector

for(bundle in (c("Lifo", "Rifo", "Lcing", "Rcing", "Fmajor","Fminor","Lilf","Rilf",
                 "Lslf","Rslf","Lunc","Runc","Fornix","Lcing2","Rcing2"))){
  colname_baselineFA <- sprintf("%s_baselineFA",bundle)
  baselineFA.baselinemotion.lm <- summary(lm(ketamine[,colname_baselineFA] ~ 
                                          ketamine$preinfusion_scan_motion + ketamine$Age))
  pvalue <- baselineFA.baselinemotion.lm$coefficients[2,4] #Obtain p-value for motion from linear model 
  p.value[bundle] <- pvalue #Save p-value into vector
}

BaselineFA_BaselineMotion_LinearModel <- data.frame(p.value)
BaselineFA_BaselineMotion_LinearModel %>% arrange(p.value)
```

*Post-infusion FA and Post-infusion in-scanner head motion*
```{r}
p.value <- double() #create empty vector

for(bundle in (c("Lifo", "Rifo", "Lcing", "Rcing", "Fmajor","Fminor","Lilf","Rilf",
                 "Lslf","Rslf","Lunc","Runc","Fornix","Lcing2","Rcing2"))){
  colname_postinfusionFA <- sprintf("%s_postinfusionFA",bundle)
  postinfusionFA.postinfusionmotion.lm <- summary(lm(ketamine[,colname_postinfusionFA] ~ 
                                          ketamine$postinfusion_scan_motion + ketamine$Age))
  pvalue <- postinfusionFA.postinfusionmotion.lm$coefficients[2,4] #Obtain p-value for motion from linear model
  p.value[bundle] <- pvalue #Save p-value into vector
}

PostInfusionFA_PostInfusionMotion_LinearModel <- data.frame(p.value)
PostInfusionFA_PostInfusionMotion_LinearModel %>% arrange(p.value)
```

*Paired T-test: Pre- and Post-Infusion Head Motion*
```{r}
t.test(ketamine$preinfusion_scan_motion, ketamine$postinfusion_scan_motion,
       paired=TRUE,alternative="two.sided")
```

### Cook's Distance

* Left CB-hippocampal portion: 24-hour HAM-D28 Percent Improvement and Baseline FA
```{r}
Lcing2.baselineFA.HAMD28change <- lm(ketamine$Lcing2_baselineFA ~ 
                                  ketamine$HDRS.24hr.percentchange + ketamine$Age) #lm
plot(Lcing2.baselineFA.HAMD28change, 5) #Residuals v. Leverage 
cooks.distance(Lcing2.baselineFA.HAMD28change) %>% sort #Cook's Distance
```

* Left SLF: 24-hour HAM-D28 Percent Improvement and Baseline FA
```{r}
Lslf.baselineFA.HAMD28change <- lm(ketamine$Lslf_baselineFA ~
                                ketamine$HDRS.24hr.percentchange + ketamine$Age) #lm
plot(Lslf.baselineFA.HAMD28change, 5)  #Residuals v. Leverage 
cooks.distance(Lslf.baselineFA.HAMD28change) %>% sort #Cook's Distance
```
* CC-forceps minor: 24-hour HAM-D28 Percent Improvement and Post-ketamine % change in FA
```{r}
Fminor.percentchangeFA.HAMD28change <- lm(ketamine$Fminor_FA_percentchange ~
                                       ketamine$HDRS.24hr.percentchange + ketamine$Age) #lm
plot(Fminor.percentchangeFA.HAMD28change, 5)  #Residuals v. Leverage 
cooks.distance(Fminor.percentchangeFA.HAMD28change) %>% sort #Cook's Distance
```

* Left UF: 24-hour HAM-D28 Percent Improvement and Post-ketamine % change in FA
```{r}
Lunc.percentchangeFA.HAMD28change <- lm(ketamine$Lunc_FA_percentchange ~
                                     ketamine$HDRS.24hr.percentchange + ketamine$Age) #lm
plot(Lunc.percentchangeFA.HAMD28change, 5)  #Residuals v. Leverage 
cooks.distance(Lunc.percentchangeFA.HAMD28change) %>% sort #Cook's Distance
```

* Right UF: 24-hour HAM-D28 Percent Improvement and Post-ketamine % change in FA
```{r}
Runc.percentchangeFA.HAMD28change <- lm(ketamine$Runc_FA_percentchange ~
                                     ketamine$HDRS.24hr.percentchange + ketamine$Age) #lm
plot(Runc.percentchangeFA.HAMD28change, 5)  #Residuals v. Leverage 
cooks.distance(Runc.percentchangeFA.HAMD28change) %>% sort #Cook's Distance
```
